Multidimensional signals and analytic flexibility: Estimating degrees of freedom in human speech analyses

### 2022-09-19 RR_manuscript/Draft_RR.Rmd
add more refs

While properties of the teams did not seem to systematically affect results, teams' estimates seem to very much depend on certain measurement choices. 
Human speech is a complex multidimensional signal.
Researchers need to make choices about what to measure, how to measure it and which temporal unit to measure it in. 
Some outcome choices seem to bias the estimates in our data in one direction while others seem to bias estimates into another. 
For example, measurements related to fundamental frequency tended to result in lower estimates while measurements related to vowel formants tended to yield higher estimates.
This asymmetry can have several causes.
First, there could be a true underlying relationship between typicality and the speech signal that manifests itself in some measures but not others and/or manifests itself negatively in one acoustic measure but positively in another.
Since we do not know the truth and our best guess based on the meta analysis suggest no relationship, this remains speculative.
Second and independent of a possible true relationship, certain measurement choices might be associated with stronger expectations relative to the research question, which might lead to stronger researcher biases.
Many researchers targeted measures related to fundamental frequency (f0), since similar functional relationships like information structure and predictability can be expressed by f0 [e.g. @grice2017integrating; @turnbull2017role]. 
Moreover, prior work as actually suggested a relationship between typicality and f0 [e.g. @dimitrova2008prosodic; @dimitrova2009did]. 
Researchers might have heard about those findings, nudging their analysis choices slightly into a particular direction.
Regardless of its cause, we have to conclude that depending on the choice of how the speech signal is operationalized, researchers might find evidence for or against a theoretically relevant effect. 
This conclusion is further supported by the fact that between-team variability was lower than between model variability. 
This is an important observation when put into context of the fact that most teams submitted many different models. 
Teams submitted up to 16 different models to test for a possible relationship between typicality and the speech signal.
The complexity of the speech signal lends itself to multiple approaches, but this plurality of hypothesis tests invites bias and can dramatically increase the rate of falsely claiming the presence of an effect [@roettger2019researcher]. 
When operating within the frequentist inferential framework, testing the same hypothesis with different dependent variables, increases the Type-I error rate if not corrected for [Tukey 1953, @benjamini1995controlling].
<!-- TR: couldnt find bibtex of tukey: Tukey, J. W. (1953). The problem of multiple comparisons. Multiple comparisons. -->
It also invites bias, possibly leading to the selective reporting of those tests that yield a desirable outcome [@kerr1998harking, @john2012measuring, @simmons2011false] while null results remain unreported [e.g., @sterling1959publication, @rosenthal1979file]. 
Fields such as the speech sciences that advance their theories based on multidimensional data should be aware of this flexibility and calibrate their confidence in empirical claims accordingly.

@@

A desired outcome of knowing that different butc reasonable measurement choices or statistical approaches might lead to  different interpretations of research data is to calibrate our (un)certainty in the strength of the collected evidence and, in turn, communicate that (un)certainty appropriately.
The fact that the choice of measurement, measurement window, and predictor choice affect the answer to the research question further suggests that research assumptions and hypotheses should be formulated in much greater detail, particularly so in regards to how measurement systems (here, the acoustic signal) and underlying conceptual constructs (here, the phonetic expression of typicality) relate to each other.
<!-- TODO: add ref to "process and statistical model" in McElreath -->
We should ideally specify the link between conceptual construct and quantitative system---the "derivation chain" [@dubin1970theory; @meehl1990summaries]---prior to data collection and analysis, including defining constructs and their relationship within the quantitative system, specifying auxiliary assumptions and boundary conditions, and defining target measurements, statistical expectations and possible (and impossible) effect magnitudes.
Without well defined derivation chains, we "are not even wrong" [@scheel2022most] because falsified expectations cannot tell us much about the conceptual constructs they are based on when the relationship between the two is underspecified.

@@

## Caveats
Our study has several limitations that need to be considered when evaluating our results. 
Importantly, there is potential for bias. 
The coordinating authors have engaged with metascientific research before and have been actively involved in methodological debates about scientific practices including transparency and statistical methods. 
We have in the past used the lack of standardized analytical approaches as an argument for proposing behavior and policy changes in the field. 
This might have biased our own judgment during the analysis which itself came with many researcher degrees of freedom. 
We hope we were able to make these degrees of freedom as well as the timing and reasoning of our analytical choices at least detectable. 
However, our results should be independently re-analysed and replicated using a different research question.
Moreover, our sample is an opportunity sample. 
We have advertised the project through online platforms which might have led to the exclusion of certain potential researcher groups.
While the number of participating teams is larger than most earlier crowdsourcing projects, it is likely to be too small to estimate for example the impact of certain predictors reliably.
Since predictor levels were not systematically distributed across teams, our estimates are characterized by large uncertainty. 
This uncertainty is possibly further inflated by the fact that he research question presented to the teams was vague: *Do speakers acoustically modify utterances to signal atypical word combinations?* 
Interpreting the research question/hypothesis differently in terms of its statistical consequences has recently been shown to explain some variation between analysis teams in many-analysis projects [@auspurg2021has]. 
However, we consider this very underspecification of research hypotheses in the field of speech science (and beyond, see @scheel2022most) a common phenomenon. 
For example, researchers seem to have not yet agreed on how to acoustically measure cross linguistically common phenomena such as word stress [e.g. @gordon2017acoustic]. 
In light of these possible sources of bias, a field can benefit from explicit positionality statements [e.g., @fox2021open]
Researchers do not analyse data in a vaccuum.
It is important recognize and disclose one’s positionality, i.e., a reflection about how educational background, social identity, power, experience and context might influence researchers approaches and interpretations.
Despite these all of these possible sources of bias and uncertainty, our study suggest that at a minimum we can expect the between-team and between-analysis variability that we have observed, which is quite substantial and possibly amplifying biases. 
Beyond the potential for bias, our study lacks generalizability. 
The generalizability of our findings to other disciplines on the one hand and specifically other sub-disciplines of the language sciences is, of course, limited.
We focused on quantitative analyses that require operationalization of a multidimensional signal. 
While some of our results can be informative for other disciplines that work with speech or video signals, they might not be very informative for expected variation between analysts working in qualitative research fields.
This is an important point to make because the cognitive sciences in general, and the language sciences in particular have many research areas that are explored with qualitative methods rather than quantitatively. 
It is conceivable that the discussed issues apply differently or not at all to qualitative data analysis.

>>>>

While properties of the teams did not seem to systematically affect results, teams' estimates seem to very much depend on certain measurement choices. 
Human speech is a complex multidimensional signal.
Researchers need to make choices about what to measure, how to measure it and which temporal unit to measure it in. 
Some outcome choices seem to bias the estimates in our data in one direction while others seem to bias estimates into another. 
For example, measurements related to fundamental frequency tended to result in lower estimates while measurements related to vowel formants tended to yield higher estimates.
This asymmetry can have several causes.
First, there could be a true underlying relationship between typicality and the speech signal that manifests itself in some measures but not others and/or manifests itself negatively in one acoustic measure but positively in another.
Since we do not know the truth and our best guess based on the meta analysis suggest no relationship, this remains speculative.
Second and independent of a possible true relationship, certain measurement choices might be associated with stronger expectations relative to the research question, which might lead to stronger researcher biases.
Many researchers targeted measures related to fundamental frequency (f0), since similar functional relationships like information structure and predictability can be expressed by f0 [e.g. @grice2017integrating; @turnbull2017role]. 
Moreover, prior work as actually suggested a relationship between typicality and f0 [e.g. @dimitrova2008prosodic; @dimitrova2009did]. 
Researchers might have heard about those findings, nudging their analysis choices slightly into a particular direction.
Regardless of its cause, we have to conclude that depending on the choice of how the speech signal is operationalized, researchers might find evidence for or against a theoretically relevant effect. 
This conclusion is further supported by the fact that between-team variability was lower than between model variability. 
This is an important observation when put into context of the fact that most teams submitted many different models. 
Teams submitted up to 16 different models to test for a possible relationship between typicality and the speech signal.
The complexity of the speech signal lends itself to multiple approaches, but this plurality of hypothesis tests invites bias and can dramatically increase the rate of falsely claiming the presence of an effect [@roettger2019researcher]. 
When operating within the frequentist inferential framework, testing the same hypothesis with different dependent variables, increases the Type-I error rate if not corrected for [@tukey1953].
<!-- TR: couldnt find bibtex of tukey: Tukey, J. W. (1953). The problem of multiple comparisons. Multiple comparisons. 
JVC: I found it as an unpublished report and added tukey1953 to the refs file. Not sure if that is better than [Tukey 1953, as cited in @benjamini1995controlling] or not.-->
It also invites bias, possibly leading to the selective reporting of those tests that yield a desirable outcome [@kerr1998harking, @john2012measuring, @simmons2011false] while null results remain unreported [e.g., @sterling1959publication, @rosenthal1979file]. 
Fields such as the speech sciences that advance their theories based on multidimensional data should be aware of this flexibility and calibrate their confidence in empirical claims accordingly.

@@

A desired outcome of knowing that different butc reasonable measurement choices or statistical approaches might lead to  different interpretations of research data is to calibrate our (un)certainty in the strength of the collected evidence and, in turn, communicate that (un)certainty appropriately.
The fact that the choice of measurement, measurement window, and predictor choice affect the answer to the research question further suggests that research assumptions and hypotheses should be formulated in much greater detail, particularly so in regards to how measurement systems (here, the acoustic signal) and underlying conceptual constructs (here, the phonetic expression of typicality) relate to each other.
<!-- TODO: add ref to "process and statistical model" in McElreath 
JVC: added @mcelreath2020 to bib file. -->
We should ideally specify the link between conceptual construct and quantitative system---the "derivation chain" [@dubin1970theory; @meehl1990summaries]---prior to data collection and analysis, including defining constructs and their relationship within the quantitative system, specifying auxiliary assumptions and boundary conditions, and defining target measurements, statistical expectations and possible (and impossible) effect magnitudes.
Without well defined derivation chains, we "are not even wrong" [@scheel2022most] because falsified expectations cannot tell us much about the conceptual constructs they are based on when the relationship between the two is underspecified.

@@

## Caveats

Our study has several limitations that need to be considered when evaluating our results. 
Importantly, there is potential for bias. 
The coordinating authors have engaged with metascientific research before and have been actively involved in methodological debates about scientific practices including transparency and statistical methods. 
We have in the past used the lack of standardized analytical approaches as an argument for proposing behavior and policy changes in the field. 
This might have biased our own judgment during the analysis which itself came with many researcher degrees of freedom. 
We hope we were able to make these degrees of freedom as well as the timing and reasoning of our analytical choices at least detectable. 
However, our results should be independently re-analysed and replicated using a different research question.
Moreover, our sample is an opportunity sample. 
We have advertised the project through online platforms which might have led to the exclusion of certain potential researcher groups.
While the number of participating teams is larger than most earlier crowdsourcing projects, it is likely to be too small to estimate for example the impact of certain predictors reliably.
Since predictor levels were not systematically distributed across teams, our estimates are characterized by large uncertainty. 
This uncertainty is possibly further inflated by the fact that he research question presented to the teams was vague: *Do speakers acoustically modify utterances to signal atypical word combinations?* 
Interpreting the research question/hypothesis differently in terms of its statistical consequences has recently been shown to explain some variation between analysis teams in many-analysis projects [@auspurg2021has]. 
However, we consider this very underspecification of research hypotheses in the field of speech science (and beyond, see @scheel2022most) a common phenomenon. 
For example, researchers seem to have not yet agreed on how to acoustically measure cross linguistically common phenomena such as word stress [e.g. @gordon2017acoustic]. 
In light of these possible sources of bias, a field can benefit from explicit positionality statements [e.g., @fox2021open]
Researchers do not analyse data in a vaccuum.
It is important recognize and disclose one’s positionality, i.e., a reflection about how educational background, social identity, power, experience and context might influence researchers approaches and interpretations.
Despite these all of these possible sources of bias and uncertainty, our study suggest that at a minimum we can expect the between-team and between-analysis variability that we have observed, which is quite substantial and possibly amplifying biases. 
Beyond the potential for bias, our study lacks generalizability. 
The generalizability of our findings to other disciplines on the one hand and specifically other sub-disciplines of the language sciences is, of course, limited.
We focused on quantitative analyses that require operationalization of a multidimensional signal. 
While some of our results can be informative for other disciplines that work with speech or video signals, they might not be very informative for expected variation between analysts working in qualitative research fields.
This is an important point to make because the cognitive sciences in general, and the language sciences in particular have many research areas that are explored with qualitative methods rather than quantitatively. 
It is conceivable that the discussed issues apply differently or not at all to qualitative data analysis.

### 2022-09-19 RR_manuscript/Draft_RR.tex
add more refs

While properties of the teams did not seem to systematically affect results, teams' estimates seem to very much depend on certain measurement choices.
Human speech is a complex multidimensional signal.
Researchers need to make choices about what to measure, how to measure it and which temporal unit to measure it in.
Some outcome choices seem to bias the estimates in our data in one direction while others seem to bias estimates into another.
For example, measurements related to fundamental frequency tended to result in lower estimates while measurements related to vowel formants tended to yield higher estimates.
This asymmetry can have several causes.
First, there could be a true underlying relationship between typicality and the speech signal that manifests itself in some measures but not others and/or manifests itself negatively in one acoustic measure but positively in another.
Since we do not know the truth and our best guess based on the meta analysis suggest no relationship, this remains speculative.
Second and independent of a possible true relationship, certain measurement choices might be associated with stronger expectations relative to the research question, which might lead to stronger researcher biases.
Many researchers targeted measures related to fundamental frequency (f0), since similar functional relationships like information structure and predictability can be expressed by f0 \citep[e.g.][]{grice2017integrating, turnbull2017role}.
Moreover, prior work as actually suggested a relationship between typicality and f0 \citep[e.g.][]{dimitrova2008prosodic, dimitrova2009did}.
Researchers might have heard about those findings, nudging their analysis choices slightly into a particular direction.
Regardless of its cause, we have to conclude that depending on the choice of how the speech signal is operationalized, researchers might find evidence for or against a theoretically relevant effect.
This conclusion is further supported by the fact that between-team variability was lower than between model variability.
This is an important observation when put into context of the fact that most teams submitted many different models.
Teams submitted up to 16 different models to test for a possible relationship between typicality and the speech signal.
The complexity of the speech signal lends itself to multiple approaches, but this plurality of hypothesis tests invites bias and can dramatically increase the rate of falsely claiming the presence of an effect \citep{roettger2019researcher}.
When operating within the frequentist inferential framework, testing the same hypothesis with different dependent variables, increases the Type-I error rate if not corrected for \citep[Tukey 1953,][]{benjamini1995controlling}.
It also invites bias, possibly leading to the selective reporting of those tests that yield a desirable outcome \citep[\citet{john2012measuring}, \citet{simmons2011false}]{kerr1998harking} while null results remain unreported \citep[e.g.,][\citet{rosenthal1979file}]{sterling1959publication}.
Fields such as the speech sciences that advance their theories based on multidimensional data should be aware of this flexibility and calibrate their confidence in empirical claims accordingly.

>>>>

While properties of the teams did not seem to systematically affect results, teams' estimates seem to very much depend on certain measurement choices.
Human speech is a complex multidimensional signal.
Researchers need to make choices about what to measure, how to measure it and which temporal unit to measure it in.
Some outcome choices seem to bias the estimates in our data in one direction while others seem to bias estimates into another.
For example, measurements related to fundamental frequency tended to result in lower estimates while measurements related to vowel formants tended to yield higher estimates.
This asymmetry can have several causes.
First, there could be a true underlying relationship between typicality and the speech signal that manifests itself in some measures but not others and/or manifests itself negatively in one acoustic measure but positively in another.
Since we do not know the truth and our best guess based on the meta analysis suggest no relationship, this remains speculative.
Second and independent of a possible true relationship, certain measurement choices might be associated with stronger expectations relative to the research question, which might lead to stronger researcher biases.
Many researchers targeted measures related to fundamental frequency (f0), since similar functional relationships like information structure and predictability can be expressed by f0 \citep[e.g.][]{grice2017integrating, turnbull2017role}.
Moreover, prior work as actually suggested a relationship between typicality and f0 \citep[e.g.][]{dimitrova2008prosodic, dimitrova2009did}.
Researchers might have heard about those findings, nudging their analysis choices slightly into a particular direction.
Regardless of its cause, we have to conclude that depending on the choice of how the speech signal is operationalized, researchers might find evidence for or against a theoretically relevant effect.
This conclusion is further supported by the fact that between-team variability was lower than between model variability.
This is an important observation when put into context of the fact that most teams submitted many different models.
Teams submitted up to 16 different models to test for a possible relationship between typicality and the speech signal.
The complexity of the speech signal lends itself to multiple approaches, but this plurality of hypothesis tests invites bias and can dramatically increase the rate of falsely claiming the presence of an effect \citep{roettger2019researcher}.
When operating within the frequentist inferential framework, testing the same hypothesis with different dependent variables, increases the Type-I error rate if not corrected for \citep{tukey1953}.
It also invites bias, possibly leading to the selective reporting of those tests that yield a desirable outcome \citep[\citet{john2012measuring}, \citet{simmons2011false}]{kerr1998harking} while null results remain unreported \citep[e.g.,][\citet{rosenthal1979file}]{sterling1959publication}.
Fields such as the speech sciences that advance their theories based on multidimensional data should be aware of this flexibility and calibrate their confidence in empirical claims accordingly.
